---
title: "Data Analysis"
author: "Shaun Jackson, Marissa Chemotti"
date: "7/3/2019"
output: 
  rmdformats::material:
    self_contained: no
---

```{r setup, include=FALSE}
options(stringsAsFactors = FALSE)
library(kableExtra)
adosm1 <- read.csv('./datasets/adosm1_scored.csv', stringsAsFactors = FALSE)
```

# Data Analysis

## Read in Data

- Our input data will be a scored version of the artificial ADOS dataset with some fake columns to demonstrate how to use R to perform specific data analyses.


```{r dataInput, include=FALSE}
adosm2 <- read.csv('./datasets/adosm2_scored.csv', 
                   stringsAsFactors = FALSE)
```

## Basic Stats functions

### Mean, Median, Standard Deviation, Summary
```{r}
mean(adosm2$ados_sarb_total, na.rm = TRUE)
median(adosm2$ados_sarb_total, na.rm = TRUE)
sd(adosm2$ados_sarb_total, na.rm = TRUE)
summary(adosm2$ados_sarb_total)
```

## Linear Regression

### Correlation

- Find correlation between two vars
```{r}
cor.test(adosm2$ados_fake_score1, adosm2$ados_fake_lin_outcome)
```

### Creating a Linear Regression Model

- We use the `lm()` functionn to create a Linear Regression Model
- The first parameter takes in an equation of the form: `dependentVar ~ predictor1 + predictor2 + ...`
    - So the dependentVar is a function of the predictor variables
- In order to use column names as variables in this eq., the `data` parameter takes in a dataframe as input.


```{r linMod}
# To demonstrate a linear model, we use fake ados columns within the ados
# dataframe
linearMod <- 
  lm(ados_fake_lin_outcome ~ ados_fake_score1 + ados_fake_score2,
     data=adosm2)
```
### Residuals, P-Values, Coefficients, with `summary()`

```{r}
summary(linearMod)
```


### Plotting the Linear Model
```{r}
# create effects object
plot(effects::Effect(c('ados_fake_score1'), linearMod))
plot(effects::Effect(c('ados_fake_score2'), linearMod))
plot(effects::Effect(c('ados_fake_score1', 'ados_fake_score2'), linearMod))
```

## Logistic Regression

```{r}
adosm2$cbe_asd <- ifelse(adosm2$cbe_36 %in% c('Autism', 'ASD'), 'ASD', 'Non-ASD')
adosm2$cbe_asd <- as.factor(adosm2$cbe_asd)
adosm2$cbe_asd <- relevel(adosm2$cbe_asd, ref = 'Non-ASD')
mod <- glm(cbe_asd ~ 1 + ados_sarb_total,
                   data = adosm2,
                   family = 'binomial')
eff <- effects::Effect(c('ados_sarb_total'), mod)
plot(eff,
     axes = list(
       y = list(
         lab = 'CBE of ASD',
         ticks = list(at = c(.01, .1, .2, .4, .6, .8,  .99))))
)


```


## Mixed Effects Models

### Mixed Effects Logistic Regression

```{r}
# analyze- logistic regression mixed effects
adosm2$cbe_asd <- ifelse(adosm2$cbe_36 %in% c('Autism', 'ASD'), 'ASD', 'Non-ASD')
adosm2$cbe_asd <- as.factor(adosm2$cbe_asd)
adosm2$cbe_asd <- relevel(adosm2$cbe_asd, ref = 'Non-ASD')
mod <- lme4::glmer(cbe_asd ~ 1 + (1|visit) + ados_sarb_total,
                   data = adosm2,
                   family = 'binomial')
eff <- effects::Effect(c('ados_sarb_total'), mod)
plot(eff,
     axes = list(
       y = list(
         lab = 'CBE of ASD',
         ticks = list(at = c(.01, .1, .2, .4, .6, .8,  .99))))
)


```


## Data Visualization with `ggplot`

### Plotting points
```{r}
library(ggplot2)
ggplot(data = adosm2) + 
  geom_point(mapping = aes(x = ados_fake_lin_outcome, 
                           y = ados_sarb_total, 
                           color = recruitment_group))

```

### Plotting with Points and Smooth

```{r}
ggplot(data = adosm2, mapping = aes(x = ados_sarb_total, y = ados_fake_lin_outcome)) + 
  geom_point(mapping = aes(color = recruitment_group)) + 
  geom_smooth()
```

### Bar Plots

#### Bar plot with Colors
```{r}
ggplot(data = adosm2) + 
  geom_bar(mapping = aes(x = recruitment_group, fill = gender))

```

#### Bar plot with separated colors and positions
```{r}
ggplot(data = adosm2) + 
  geom_bar(mapping = aes(x = recruitment_group, fill = gender), position = "dodge")
```

### Box Plots

```{r}
ggplot(data = adosm2, mapping = aes(x = recruitment_group, y = ados_sarb_total)) +
  geom_boxplot()
```

## More Resources

