---
title: "Data Analysis"
author: "Shaun Jackson, Marissa Chemotti"
date: "`r Sys.Date()`"
output: 
  rmdformats::material:
    self_contained: no
---

```{r setup, include=FALSE}
options(stringsAsFactors = FALSE)
library(kableExtra)
```

# Data Visualization

## Read in Data

- Our input data will be a scored version of the artificial ADOS Module 2dataset with some fake columns to demonstrate how to use R to perform specific data analyses.


```{r dataInput}
adosm2 <- read.csv('./datasets/adosm2_scored.csv', 
                   stringsAsFactors = FALSE)
```

## Data Visualization with `ggplot`

### Plotting points
```{r}
library(ggplot2)
ggplot(data = adosm2) + 
  geom_point(mapping = aes(x = ados_fake_lin_outcome, 
                           y = ados_sarb_total, 
                           color = recruitment_group))

```

### Plotting with Points and Smooth

```{r}
ggplot(data = adosm2, mapping = aes(x = ados_sarb_total, y = ados_fake_lin_outcome)) + 
  geom_point(mapping = aes(color = recruitment_group)) + 
  geom_smooth()
```

### Bar Plots

#### Bar plot with Colors
```{r}
ggplot(data = adosm2) + 
  geom_bar(mapping = aes(x = recruitment_group, fill = gender))

```

#### Bar plot with separated colors and positions
```{r}
ggplot(data = adosm2) + 
  geom_bar(mapping = aes(x = recruitment_group, fill = gender), position = "dodge")
```

### Box Plots

```{r}
ggplot(data = adosm2, mapping = aes(x = recruitment_group, y = ados_sarb_total)) +
  geom_boxplot()
```

