---
title: "Data Analysis"
author: "Shaun Jackson, Marissa Chemotti"
date: "`r Sys.Date()`"
output: 
  rmdformats::material:
    self_contained: no
---

```{r setup, include=FALSE}
options(stringsAsFactors = FALSE)
library(kableExtra)
```


# Data Visualization

- We will use the ggplot2 library for data visualization

```{r}
library("ggplot2")
# if not installed:
# install.packages("ggplot2")
```

## Read in Data

- Our input data will be a scored version of the artificial ADOS Module 2dataset with some fake columns to demonstrate how to use R to perform specific data analyses.


```{r dataInput}
adosm2 <- read.csv('./datasets/adosm2_scored.csv', 
                   stringsAsFactors = FALSE)
```

## Our dataset
```{r}
head(adosm2)
```

## Data Visualization with `ggplot`

### Resources

- This lesson largely borrowed and used the material in this book:
- [R for Data Science](https://r4ds.had.co.nz/)

- Explore Chapter 3 of the book for a much more in depth lesson

### Plotting points
```{r}
library("ggplot2")
ggplot(data = adosm2) + 
  geom_point(mapping = aes(x = ados_fake_lin_outcome, 
                           y = ados_sarb_total, 
                           color = recruitment_group))

```

- `ggplot()` will create a coordinate system you can add layers on top of
    - it takes in the dataset to use in the graph with the `data` parameter
- `geom_point()` is the extra layer we are adding on the coordinate system, so we are adding points to it
    - the `mapping` argument will define how variables in your data will be **mapped** to a visual property
        - The `aes()` argument within marks what piece of your dataset will map to the x or y axes, or even colors or fills.

### Plotting with Points and Smooth

```{r}
ggplot(data = adosm2, mapping = aes(x = ados_sarb_total, y = ados_fake_lin_outcome)) + 
  geom_point(mapping = aes(color = recruitment_group)) + 
  geom_smooth()
```

- We can plot points  with a smooth line by adding another layer with `+ geom_smooth()`.
- In this example, the first layer takes in our adosm2 dataset, and maps the ados_sarb_total to the x axes,and the ados_fake_lin_outcome to the y axes
- The second layer plots points using the mapping of the first layer, but in addition adds color to the points by recruitment group
- The third layer uses the mapping of the first layer, so does not need any additional arguments

- We can similarly create bar plots and box plots using the same syntax, but with `geom_bar()` and `geom_boxplot()`


### Bar Plots

#### Bar plot with Colors
```{r}
ggplot(data = adosm2) + 
  geom_bar(mapping = aes(x = recruitment_group, fill = gender))

```

#### Bar plot with separated colors and positions
```{r}
ggplot(data = adosm2) + 
  geom_bar(mapping = aes(x = recruitment_group, fill = gender), position = "dodge")
```

### Box Plots

```{r}
ggplot(data = adosm2, mapping = aes(x = recruitment_group, y = ados_sarb_total)) +
  geom_boxplot()
```

## Template for graphing

```{text}
ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))
```

- You can search what other geom function are available by following the links on your help page:

- `?ggplot2`