---
title: "04-Data-Analysis"
author: "Shaun Jackson, Marissa Chemotti"
date: "`r Sys.Date()`"
output: 
  rmdformats::material:
    self_contained: no
---

# Data Analysis

## Load required libraries
```{r loadLibraries}
library("lme4")
library("effects")
```

## Read in Data

- Our input data will be a scored version of the artificial ADOS Module 2dataset with some fake columns to demonstrate how to use R to perform specific data analyses.


```{r dataInput}
adosm2 <- read.csv('./datasets/adosm2_scored.csv', 
                   stringsAsFactors = FALSE)
```


## Basic Stats functions

### Mean, Median, Standard Deviation, Summary
```{r}
mean(adosm2$ados_sarb_total, na.rm = TRUE)
median(adosm2$ados_sarb_total, na.rm = TRUE)
sd(adosm2$ados_sarb_total, na.rm = TRUE)
summary(adosm2$ados_sarb_total)
```

### Chi Squared Test

```{r}
chisq.test(table(adosm2$cbe_36, adosm2$recruitment_group))
```

## Linear Regression

### Correlation

- Find correlation between two vars
```{r}
cor.test(adosm2$ados_fake_score1, adosm2$ados_fake_lin_outcome)
```

### Creating a Linear Regression Model

- We use the `lm()` functionn to create a Linear Regression Model
- The first parameter takes in an equation of the form: `dependentVar ~ predictor1 + predictor2 + ...`
    - So the dependentVar is a function of the predictor variables
- In order to use column names as variables in this eq., the `data` parameter takes in a dataframe as input.


```{r linMod}
# To demonstrate a linear model, we use fake ados columns within the ados
# dataframe
linearMod <- 
  lm(ados_fake_lin_outcome ~ ados_fake_score1 + ados_fake_score2,
     data=adosm2)
```
### Residuals, P-Values, Coefficients, with `summary()`

```{r}
summary(linearMod)
```


### Plotting the Linear Model
```{r}
# create effects object
plot(Effect(c('ados_fake_score1'), linearMod))
plot(Effect(c('ados_fake_score2'), linearMod))
plot(Effect(c('ados_fake_score1', 'ados_fake_score2'), linearMod))
```

## Logistic Regression

```{r}
# We must dichotomize our dependent variable
# We will create a model that predicts a clinical estimate of ASD
adosm2$cbe_asd <- ifelse(adosm2$cbe_36 %in% c('Autism', 'ASD'), 'ASD', 'Non-ASD')
# Now there are only two values: ASD, Non-ASD
# We now convert this into a factor, so it can
# be used to create a model
adosm2$cbe_asd <- as.factor(adosm2$cbe_asd)
# We define what our reference level will be
# So Non-ASD will be at 0, and ASD will be 1
adosm2$cbe_asd <- relevel(adosm2$cbe_asd, ref = 'Non-ASD')
# glm takes in an equation, dataset, and family
# for what model to create
mod <- glm(cbe_asd ~ 1 + ados_sarb_total,
                   data = adosm2,
                   family = 'binomial')
# Create an effects object for plotting
eff <- Effect(c('ados_sarb_total'), mod)
# View values
eff
# Create plot with axes labels
plot(eff,
     axes = list(
       y = list(
         lab = 'CBE of ASD',
         ticks = list(at = c(.01, .1, .2, .4, .6, .8,  .99))))
)


```


## Mixed Effects Models

### Mixed Effects Logistic Regression

```{r}
# analyze- logistic regression mixed effects
adosm2$cbe_asd <- ifelse(adosm2$cbe_36 %in% c('Autism', 'ASD'), 'ASD', 'Non-ASD')
adosm2$cbe_asd <- as.factor(adosm2$cbe_asd)
adosm2$cbe_asd <- relevel(adosm2$cbe_asd, ref = 'Non-ASD')
mod <- glmer(cbe_asd ~ 1 + (1|visit) + ados_sarb_total,
                   data = adosm2,
                   family = 'binomial')
eff <- Effect(c('ados_sarb_total'), mod)
plot(eff,
     axes = list(
       y = list(
         lab = 'CBE of ASD',
         ticks = list(at = c(.01, .1, .2, .4, .6, .8,  .99))))
)


```

