---
title: "Intro to R- Part 1: Data Exploration"
author: "Shaun Jackson, Marissa Chemotti"
date: "7/3/2019"
output: 
  html_document:
    theme: journal
    toc: true
    toc_depth: 4
    toc_float: true
---

```{r setup, include=FALSE}
options(stringsAsFactors = FALSE)
library(kableExtra)
```

<div style="margin-bottom:100px;">
# Introduction
  <hr />
  
The aim of this document is to present you with the bare minimum essentials to get you up and running with R.
One of the toughest parts about  learning a programming language is feeling overwhelmed by the abundance of material to learn.

By giving you an overview of the most useful functions, in the lens of a real dataset, we hope that you can apply these concepts to your own data.

As with everything else, the best way to learn R is through **consistent** practice with a real world project.

<hr />
## Goals of this Document
By the end of this document, you will be able to:

1. Learn about the fundamental R Datatypes
2. Read a CSV file into your R environment
3. Write from your R environment to a CSV file
4. Explore different attributes of a dataframe
    - column names
    - row names
    - more with `attributes()`
    - Subset a dataframe by rows and columns
    - Filter a dataframe by the values of a column
5. Use two data exploration and cleaning functions
    - Frequency counts of unique values with `table()`
    - Search for any duplicates within your data with `duplicated()`
  
</div>

<div style="margin-bottom:100px;">
# Datatypes
  <hr />

## Numeric, Logical, Character, Factor

## Lists, Vectors, Matrices

## Dataframes
</div>

<div style="margin-bottom:100px;">
# Reading in Data
  <hr />

## Read CSV Files

- store the filepath within a character variable
- the filepath could be relative to your working directory, or absolute
- `read.csv` is a function that could read CSV files into your R environment

```{r readCSV, message=F, warning=F}

# define relative filepath
pathToAdosFile <- './datasets/adosm1.csv'

# Use read.csv() and filepath variable as first argument to function
adosm1 <- read.csv(pathToAdosFile, 
                   stringsAsFactors = FALSE)
# stringsAsFactors is an optional argument that tells read.csv to
# not turn a string column into a R factor variable

# adosm1 is now considered a datraframe

```

## `View()`

- To view your dataframe call the `View()` function.
- Note the capital **V**iew().

```{r view, message=F, warning=F}
View(adosm1)
```

## See Dataframe attributes

- display column names `names()`
- display rownames with `row.names()`
- see all attributes with `attributes(dataframe)`

```{r }
# display column names
# names(dataframe), we will use our adosm1 dataframe
# wrap function with head() to show first 5
head(names(adosm1))

# display rownames, wrap function with head() to show first 5
head(row.names(adosm1))
```
</div>

<div style="margin-bottom:100px;">
# Writing Data
  <hr />

## `write.csv`
- We can write data from our R environment to a CSV file with `write.csv`
- `write.csv` takes in a dataframe within our R environment, as the first parameter
- the second parameter is the location and filename to write it to.

```{r writeCSVEx}
# write.csv(adosm1, 'datasets/practice write csv ados.csv', row.names = FALSE)
```
  
## `data.table::fwrite()`
- One you are comfortable with write.csv, I would suggest looking into the `fwrite` function within data.table. It is much much faster and more efficient.
  
</div>

<div style="margin-bottom:100px;">
# Subsetting Data
  <hr />
  
## Subset by columns

- In our case, adosm1 is within our R environment, so we would call
    - `adosm1[namesOfRows, namesOfColumns]`
    - Note: it is best practice to access columns by their names, not their numbers

```{r subsetCols, results="hide"}
# To access the id column, we would write
adosm1[, "id"]

# To access id and visit
adosm1[, c("id", "visit")]

# in order to make code less verbose, when accessing one column, we use
# the $ operator
# to access the id column
adosm1$id

```

## Subset by rows

- We could similarly access rows by rownames or row numbers
     - `adosm1[c(1, 2, 5), ]`
     
```{r subsetRows, results="hide"}
# access 1st, 2nd, and 5th rows
adosm1[c(1, 2, 5), ]

```

## Subset by rows and columns

```{r subsetRowsCols, results="hide"}
# access 1st, 2nd, and 5th rows, within id and visit columns
adosm1[c(1, 2, 5), c("id", "visit")]
```
</div>


<div style="margin-bottom:100px;">
# Filtering a dataframe: Subsetting part 2
  <hr />
  
## Subset with a logical vector


- What is very powerful in this is we could subset a dataframe in order to answer specific questions
- This is done by using a **logical vector** to subset or slice a dataframe

## Subset logical vector example


```{r subsLogical, results="hide"}
# To Filter for all visits at 36 months:
# Create logical vector
visits36mo <- adosm1$visit == "36"

# use  our visits36mo  logical vector to filter dataframe
adosm1[visits36mo, ]  # returns  a dataframe of all visits at 36mo

adosm1[visits36mo, "id"] # returns all the ids with a visit at 36mo

# returns alg and version at 36 months
adosm1[visits36mo, c("ados_algorithm", "ados_version")] 

```
</div>

<div style="margin-bottom:100px;">
# Functions to explore data

- Out of all the functions to  explore a dataframe, I  personally find 
    - `table()`
    - `duplicated()`
    
to be the most useful

## Example `table()`

```{r EDA}

# To get a frequency count of  all unique values,
# we use the  table()

# table() takes in 1 or more vectors as input
# since every dataframe column could be considered a vector (or list)
# we could check for every unique visit with freq counts with
table(adosm1$visit)
# we could check  for every unique combination of vectors by 
# providing multiple vector inputs

# To see every unique combination of visit and ados algorithm
table(adosm1$visit, adosm1$ados_algorithm)
```


## Checking for duplicates: `duplicated()`
- We will first generate a dataframe that has duplicates, to show how to detect duplicates by different subgroups


```{r duplicatesCheck1}

dupDf <- data.frame(id = c(1, 1, 2, 2, 3, 3),
                    visit = c('10 months', '20 months',  '10 months', '10 months', '20 months', '20 months'),
                    item1 = c(1, 2, 3, 3, 5, 6))


```

## `duplicated()` explained
- The `duplicated` function returns a logical vector of the duplicates that exist in a dataframe.
- The TRUE values represent the index locations of where a duplicate row was found.
- We could use this logical vector to see where a duplicate exists

## `duplicated()` example
```{r duplicatesCheck2}

duplicateIndices <- duplicated(dupDf)  # store dup vector in variable

duplicateIndices  # view logical vector of duplicates

dupDf[duplicateIndices, ]
```

## `Duplicated()` by key variables: id and visit

- we could check for duplicates by columns that should represent a unique combination.
- in this example: id and visit

```{r duplicatedCheck3}
# to check duplicates by groups
dupIndGrouped <- duplicated(dupDf[, c("id", "visit")])
dupDf[dupIndGrouped, ]
```
</div>

<div style="margin-bottom:100px;">
# Assignment: adosm2 dataset
  <hr />

## Instructions

1. Read in the adosm2.csv file
2. What is the frequency count for every unique visit?
3. What are the ids that exist for the 30 month visit?
4. Are there duplicates by id and visit?
5. What is the frequency count of duplicates vs non-duplicates?

</div>

<div style="margin-bottom:300px">
# Additional Resources
<hr />
</div>

<div style="margin-bottom:100px">
# Solution
<hr />

```{r assignmentSolutions}
# 1. Read in the adosm2.csv file
adosm2Path <- './datasets/adosm2.csv'
adosm2 <- read.csv(adosm2Path, stringsAsFactors = FALSE)

# 2. What is the frequency count for every unique visit?
table(adosm2$visit)

# 3, What are the ids that exist for the 30 month visit?
adosm2[adosm2$visit == "30", "id"]
# This works too- adosm2$id[adosm2$visit == "30"]

# 4. Are there duplicates by id and visit?
any(duplicated(adosm2[, c("id", "visit")]))

# 5. What are the frequency count of duplicates vs non-duplicates
table(duplicated(adosm2))




```

</div>

# Bonus Material (Optional)

## String Manipulation

- What are the most useful use cases for string manipulation?
    - Pattern matching: One can easily automate trudging through a dataset by hand, by finding patterns within the character columns in their dataset
    - Data cleaning, Are their some mispelled or case sensitive datasets? How do we group these cases?

### Regular Expressions

- To find patterns within data that has characters within it, we use **regular expressions**

Regular expressions are expressions that allow one to find patterns. 

- To implement a regular expression in R, we can use the `grep` function to find patterns, or `gsub` to both find patterns and substitute patterns with another one.

#### grep

- `grep` is a function that we will use to find patterns in a vector of strings
- the first argument will be the regular expression
- the second argument will be the vector we are searching within

```{r grepExample}
stringVecToSearch <- c('a', 'b', 'c', 'd', 'bad', 'jab')
regExprPattern <- 'a'
grep(regExprPattern, stringVecToSearch)  # indices of all strings with  an 'a'

# Find patterns that STARTS with a pattern with ^
regExprPattern <- '^b'
# in this case, starts with a 'b'
grep(regExprPattern, stringVecToSearch)

# Find patterns that ENDS with a pattern with $
regExprPattern <- 'b$'
# in this case, ends with a b
grep(regExprPattern, stringVecToSearch)

# If we want to see the values, not the indices of where patterns exist
grep(regExprPattern, stringVecToSearch, value=TRUE)
```

#### Grep Applied to a dataframe

```{r grepApplied, results=FALSE}
# see all columns that STARTS with ados
ados_cols <- grep('^ados', names(adosm1), value=TRUE)
adosm1[, ados_cols]

# see all rows that have a an ados algorithm of no words
# grep will return row numbers of where a pattern of no words
# exists in ados_version
no_words_vals <- grep('no words', adosm1$ados_version)
# use row numbers to filter dataframe
adosm1[no_words_vals, ]
```



#### gsub

- gsub can be thought of as using grep to find the patterns, but replacing the patterns found with something new.
- The first argument is the pattern to find
- The second argument is what to replace with the pattern found
- The third argument is the vector that holds the pattern to search and replace

```{r gsubEx}
stringVecToSub <- c('id', 'visit', 'ad1', 'ad2', 'ad3', 'ad4')
gsub('ad', 'ados_', stringVecToSub)

```

### String Concatenation

#### paste

#### paste0

#### sprintf

## DataFrame Manipulation Extras

### Rename columns

## Programming Constructs

### control statements essentials

### if else

### for loops


## Vectorization

### ifelse()

## Write Clean Code

### Max Line Length

### Speak through Code

### A Comment on Comments







